---
title: "picaplot Vignette"
author: "Jin Hyun Ju"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `picaplot` package provides functions that perform principal or independent component analysis and easy visualization of the results. 

## Quick Start

### Important Note

The report generating feature of this package requires a program called ```pandoc``` version 1.12.3 or higher to be installed. This is not a problem when you are using the most recent version of Rstudio (1.0.136 at the moment - 16 January 2017), since it comes with the required ```pandoc``` functionality. However, if you are running ```picaplot``` from good-old-fashioned-R (whether you are running it on a cluster or on your local machine) you might run into an error message that pandoc is not installed. In such a case, please review the following link to install the correct version of pandoc on your machine. (https://github.com/rstudio/rmarkdown/blob/master/PANDOC.md)

### 1.Loading and Preparing an Example Dataset

If you already know what you want to use the package for, follow this simple example to get started!

#### 1) Installing the package through the function ```install_github``` from the package ```devtools```.

- `picaplot` has a few dependencies and you need to install them manually before you install `picaplot`. Simply run the code below to install the dependencies.

```{r dependencies, eval = FALSE}

picaplot_dependencies <- c("ggplot2", "knitr", "rmarkdown", "devtools")

install.packages(picaplot_dependencies)

``` 

- To install `picaplot` from github:

```{r install, eval = FALSE}

devtools::install_github("jinhyunju/picaplot") #installing picaplot

```

#### 2) Loading an example dataset

Here we are going to use a public dataset that is available on the Gene Expression Omnibus (GSE60028). 

-Dhingra N, Shemer A, Correa da Rosa J, Rozenblit M et al. Molecular profiling of contact dermatitis skin identifies allergen-dependent differences in immune response. J Allergy Clin Immunol 2014 Aug;134(2):362-72. PMID: 24768652

You can also start with using your own dataset, it just needs to be a matrix which has the dimension of (gene x samples). A dataframe with covariate information is optional with dimensions (samples x covariates). The data is included with the package and can be loaded into the environment by simply using the `data()` function.


```{r load data}
library(picaplot)

data(expr_data, sample_info, probe_info)

```

To generate the example dataset yourself, you can source the script included in the package. 

- Note: If you are interested in the details of the script you can check the path to the script by printing out the value that is saved in the ```example.data.script``` object and open it up in any text editor.


```{r example_data_script, eval = FALSE}
example.data.script <- system.file("templates/create_example_data.R", package="picaplot")

source(example.data.script)
```

- Please be aware that the script will install two packages ```GEOquery``` and ```biomaRt``` if you don't already have it on your machine. The process will take a few minutes depending on your internet connection, since it is downloading data from GEO and biomaRt. If everything ran correctly, it will generate 3 objects, ```expr_data```, ```sample_info```, and ```probe_info```.

- ```expr_data``` = gene expression measurements for 26391 probes and 47 samples. 

- ```sample_info``` = 5 covariates for each sample 

- ```probe_info``` = positional information for 26391 probes
  
One thing that you have to watch out for is that the rownames of ```sample_info``` have to match the column names of the ```expr_data```. They don't necessarily have to be in the same order but they should have names that overlap. 

### 2. Core Functionality of the package

#### 1) Running PCA / ICA 

The functions for running PCA / ICA on an expression matrix are `run_pca()` and `run_ica()` respectively.  

```{r set_seed, echo = FALSE}

set.seed(1987)

```


```{r run ica and pca}
# run PCA 
pca_object <- run_pca(expr_data)

# run ICA
ica_object <- run_ica(expr_data)

```

This generates a PCAobject / ICAobject with the outputs saved in the format of a list.

#### 1-1) `run_pca()` outputs

The following entries will be generated in the output list `pca_object` after running the example above. 

  * `rotation` : Matrix of principal component gene weights where each column represents a single component. (standard `prcomp()` output)
  
  * `x` : Matrix of the projections of the original data onto principal componets. Each column holds a projection. (standard `prcomp()` output)
  
  * `sdev` : The standard deviation (square root of the eigen values) of each principal components. (standard `prcomp()` output)
  
  * `percent_var` : The percent variance each principal component is explaining. Calculated based on `sdev`
  
  * `peaks` : Indicating which gene has a gene weight larger than 2 standard deviations of its component gene weights. 
  
  * `center` : The mean values for each gene used to center the data. (standard `prcomp()` output)
  
  * `scale` : TRUE or FALSE value indicating whether the data was scaled. (standard `prcomp()` output)
  
  * Three attributes are set within the list object. "PCAobject" for `class`, "pca" for `method` and "no" for `covar_cor`.

#### 1-2) `run_ica()` outputs

The following entries will be generated in the output list `ica_object` after running the example above. 

  * `A` : The IC coefficient matrix, with each row representing coefficients for the corresponding independent component. (standard `fastICA()` output)
  
  * `S` : Matrix of gene weights for each independent component. Each column holds a single component. (standard `fastICA()` output)

  * `percent_var` : The percent variance each independent component is explaining.
  
  * `peaks` : Indicating which gene has a gene weight larger than 2 standard deviations of its component gene weights. 
  
  * `order` : The order of independent components based on the variance that they explain. 
  
  * `X`, `K`, `W` : Standard outputs of `fastICA()`. `X` is the pre-processed data matrix, `K` is the pre-whitening matrix projecting the data onto the first n principal components, and `W` is the estimated unmixing matrix. 
  
  * Three attributes are set within the list object. "ICAobject" for `class`, "ica" for `method` and "no" for `covar_cor`.

#### 2) Testing Associations Between Covariates and Components

As an optional step, you can check whether any covariates are associated with any of the components with the function `covar_association_check()` that can be applied to both PCA and ICAobjects. In this example, we are going to test the associations between the covariats in `sample_info` and each PC and IC. 

```{r covar_association_check}

pca_object <- covar_association_check(pca_object, 
                                      covars = sample_info)

ica_object <- covar_association_check(ica_object, 
                                      covars = sample_info)

```

This will add the following entries to the list.

  * `covar_pvals` : A matrix of p-values with the dimension of number of components x tested covariates. 
  
  * `comp_cov` : A list with length equal to the number of components that shows in each entry which covariates have a p-value lower than the set threshold. 
  
  * `covars` : A copy of the supplied `sample_info` for plotting. 
  
  * `covar_threshold` : The threshold for calling a covariate association significant. The default is set to 0.05 divided by the number of tests (= `length(covar_pvals)`).


#### 3) Applying unsupervised clustering to IC coefficients / PC projections

To identify any sample clusters that are not associated with measured covariates, `picaplot` provides an optional unsupervised clustering approach using functionalities from the `mclust` package. If both covariate association testing and clustering information are available, the plotting function will attempt to use the associated covariates first and if they are not available use clustering labels generated by `mclust` for each component. 

```{r detect_clusters}

ica_object <- detect_clusters(ica_object)

pca_object <- detect_clusters(pca_object)

```

#### 4) Plotting Individual Components

Individual components can be inspected by using the `plot_component()` function. This will generate 3 plots showing the gene loading on the component of interest and the component coefficients. You can specify which component to inspect by setting the component index in the option `comp_idx`.

```{r plot_indv_comp, fig.align='center', fig.width= 7, fig.height = 3}

pca1_plot = plot_component(pca_object, 
                           comp_idx = 1)

ica1_plot = plot_component(ica_object,
                           comp_idx = 1)

```

If you have information regarding the chromosome and position of each gene you can supply it to the function to color the gene loading plot by chromosome. This `geneinfo_df` dataframe needs the following columns: `phenotype` which has an entry for all rownames of the `expr_data`, 
`pheno_chr` showing which chromosome the corresponding gene is on, `pheno_start` for the starting base position of the given phenotype, 
`pheno_end` for the end base position of the phenotype. 

- Note that for ideal results you want to have the `geneinfo_df` sorted by chromosome and position. 
`plot_component` uses the order of `phenotype` given in the dataframe to generate the plot, so the user has control over how to plot it by changing the order of the `geneinfo_df` dataframe.  

```{r plot_comp_with_gene_info, fig.width= 7, fig.height = 3}

pca1_color = plot_component(pca_object, 
                            comp_idx = 1, 
                            geneinfo_df = probe_info)

ica1_color = plot_component(ica_object, 
                            comp_idx = 1, 
                            geneinfo_df = probe_info)

```

#### 5) Generate a Report for All Components

To create an HTML report showing all components with more detail use the `reportgen()` function. You can control the number of components to be plotted by setting `n_comps`, if `n_comps` is not specified it will plot every component. The default order is set by the amount of variance each component explains. If option `output_path` is not set, it will generate the report and plots in the current working directory.

```{r report_gen, eval = FALSE}

reportgen(pca_object,  prefix = "PCAreport", geneinfo_df = probe_info)

reportgen(ica_object,  prefix = "ICAreport", geneinfo_df = probe_info)


```

#### 6) Generate a Covariate Matrix for a Linear Model

After testing for covariate associations and clustering the user can generate a covariate matrix of IC coefficients or PC projections using the `get_covariate_mx` function. By default it will return a matrix that contains IC coefficients or PC projections of components associated with known covariates or those with multiple clusters detected. The user can specify the index of the components to customize the matrix in the option `idx`.


```{r create_covar_mx}

ic_covar_mx = get_covariate_mx(ica_object)

pc_covar_mx = get_covariate_mx(pca_object)

```
